---
title: "final analysis"
author: "Max Z. Y. J. Chien"
date: "2024-10-28"
output: html_document
---

```{r libraries}
library(tidyverse)
```

## Load data in

### Nanopore sequencing data

```{r}
for_final_analysis <- read_csv("../9_analyse results from minION with Flow Cell (R10.4.1)/for_final_analysis.csv.gz") %>% 
  mutate(barcode = sub("_____.*", "", insert)) %>% 
  mutate(insert = sub(".*_____", "", insert))
```

### oPools

-   Each insert is assembled by three oPools: oPools_1, oPools_2, and oPools_3

-   "UG_amount" means how many UGs are in the insert sequence

-   I labelled each insert like this "wild_type---0%\_#1---0%\_#1", each oPools separated by "---"

-   Both of "0%\_#1" and "10%\_25%\_#1" are wild type sequence, so manually change "10%\_25%\_#1" to "0%\_#1"

-   Additionally, I also calculate the UG amount of "first 150 bp of cryptic 3 prime UTR"

```{r}
potential_combination_of_inserts <- read_csv("../6_potential combination after Gibson assembly/CSV_file_of_potential_combination_of_inserts.csv") %>% 
  
  mutate(insert = str_replace(insert, "10%_25%_#1", "0%_#1")) %>% 
  
  mutate(oPools_1 = word(insert, 1, sep='---')) %>% 
  mutate(oPools_2 = word(insert, 2, sep='---')) %>% 
  mutate(oPools_3 = word(insert, 3, sep='---')) %>% 
  
  mutate(oPools_1_Seq = paste0("CAAAAGACTCCTGGGCTCACCTGTTAGC", str_extract(sequence, paste0("(?<=", "CAAAAGACTCCTGGGCTCACCTGTTAGC", ").*?(?=", "TCCTGTAAATATATATTTTTCAAAGTGA", ")")), "TCCTGTAAATATATATTTTTCAAAGTGA")) %>% 
  mutate(oPools_2_Seq = paste0("TCCTGTAAATATATATTTTTCAAAGTGA", str_extract(sequence, paste0("(?<=", "TCCTGTAAATATATATTTTTCAAAGTGA", ").*?(?=", "AAACAATAGTAGGGACTGCATATCGGTA", ")")), "AAACAATAGTAGGGACTGCATATCGGTA")) %>% 
  mutate(oPools_3_Seq = paste0("AAACAATAGTAGGGACTGCATATCGGTA", str_extract(sequence, paste0("(?<=", "AAACAATAGTAGGGACTGCATATCGGTA", ").*?(?=", "ACGAGAAGGGAACGGGGACTGCAGCCCT", ")")), "ACGAGAAGGGAACGGGGACTGCAGCCCT")) %>% 
  
  mutate(first_150_bp_of_cryptic_3_prime_UTR = paste0(str_extract(sequence, paste0("(?<=", "TCCTGTAAATATATATTTTTCAAAGTGA", ").*?(?=", "AAACAATAGTAGGGACTGCATATCGGTA", ")")), "AAACAATAGTAGGGACTGCATA")) %>% 
  
  mutate(oPools_1_UG_amount = str_count(oPools_1_Seq, "TG")) %>% 
  mutate(oPools_2_UG_amount = str_count(oPools_2_Seq, "TG")) %>% 
  mutate(oPools_3_UG_amount = str_count(oPools_3_Seq, "TG")) %>% 
  mutate(first_150_bp_of_cryptic_3_prime_UTR_UG_amount = str_count(first_150_bp_of_cryptic_3_prime_UTR, "TG")) %>% 
  
  select(-sequence, -oPools_1_Seq, -oPools_2_Seq, -oPools_3_Seq, -first_150_bp_of_cryptic_3_prime_UTR) %>% 
  
  # re-label to make everything clear
    # I know it's confusing
    # when I was designing the sequences, 200% meant that I put more UG into the sequence instead of removing it
    # at the end, total amount of UG in 200% is two times of wild type, soo it's 100% more than wild type
    # let's change it to +100%
  mutate(oPools_2 = ifelse(str_detect(oPools_2, "200%"), "+100%_1", paste0("-", oPools_2))) %>% 
  mutate(oPools_3 = ifelse(str_detect(oPools_3, "200%"), "+100%_1", paste0("-", oPools_3)))
```

## Start processing

### Quality control

check the read length

```{r}
df_1 <- for_final_analysis %>% 
  filter(!is.na(PAS_in_reference)) %>% 
  
  # if it was using canonical PAS, the read must be between 1300 and 1550 bp long
  # if it was using cryptic PAS, the read must be between 2350 and 2600 bp long
  filter(read_length >= 1300 & read_length <= 2600) %>% 
  filter(read_length <= 1550 | read_length >= 2350) %>% 
  
  group_by(insert, PAS_in_reference, contains_intron, barcode) %>% 
  mutate(new_read_counts = sum(read_counts)) %>%
  select(-read_length, -read_counts) %>% 
  unique()
```

check reads contain introns

```{r}
intron_df <- for_final_analysis %>% 
  group_by(insert, contains_intron) %>% 
  mutate(total_reads = sum(read_counts)) %>% 
  ungroup() %>% 
  distinct(insert, contains_intron, total_reads) %>% 
  pivot_wider(names_from = contains_intron, values_from = total_reads, values_fill = 0) %>% 
  mutate(total = `TRUE` + `FALSE`) %>% 
  mutate(intron_fraction = `TRUE` / total) %>% 
  arrange(desc(intron_fraction))

ggplot(intron_df) + 
  geom_col(aes(insert, intron_fraction))

inserts_with_intron <- intron_df %>% 
  filter(intron_fraction > 0.10) %>% 
  filter(total > 1000) %>% # read amount needs to be more than 1000
  pull(insert)
```

### Determine the APA site

```{r}
df_2 <- df_1 %>% 
  # get rid of those that contain introns
  filter(contains_intron == FALSE) %>% 
  filter(!insert %in% inserts_with_intron) %>% 
  
  mutate(APA_site = ifelse(PAS_in_reference > 2000, ifelse(PAS_in_reference < 2100, "canonical", "other"), "other")) %>% 
  mutate(APA_site = ifelse(PAS_in_reference > 3025, ifelse(PAS_in_reference < 3125, "cryptic", APA_site), APA_site))
```

### Calculate the fraction of cryptic PAS usage

```{r}
df_3 <- df_2 %>% 
  group_by(insert, barcode, APA_site) %>% 
  mutate(counts = sum(new_read_counts)) %>% 
  ungroup() %>% 
  select(insert, barcode, APA_site, counts) %>% 
  unique() %>% 
  pivot_wider(names_from = APA_site, values_from = counts, values_fill = 0) %>% 
  mutate(cryptic_fraction = cryptic / (canonical + cryptic + other)) %>% 
  
  # extra quality control, make sure each got at least more than 1 read
  filter(canonical > 1 & cryptic > 1)
```

### Finalise the data frame

```{r}
df_4 <- df_3 %>% 
  # turn barcode into the doxycycline (Dox) treatment group and replicate
  mutate(barcode = as.numeric(str_replace(barcode, 'barcode', ''))) %>% 
  mutate(Dox_Conc. = case_when(barcode %in% 1:3 ~ 1000,
                         barcode %in% 4:6 ~ 60,
                         barcode %in% 7:9 ~ 30,
                         barcode %in% 10:12 ~ 0)) %>% 
  mutate(replicate = barcode %% 3) %>% 
  mutate(replicate = ifelse(replicate == 0, 3, replicate)) %>% 
  select(-barcode) %>% 
  
  mutate(insert = str_replace(insert, "10%_25%_#1", "0%_#1")) %>% 
  left_join(potential_combination_of_inserts, by = "insert")
```

## Make plots

### Main (for the paper publication)

-   When I designed oPools sequences, I forgot to add wild type sequence in oPools_2

-   In order to avoid **covariance**, let's focus on the data which only involve wild type sequence in oPools_1 and oPools_3

-   This means that the data, which we are going to plot, only has UG amount change in oPools_2

-   Moreover, the first 28 bp of oPools_2 is overlapping with oPools_1, which is also in the canonical 3'-UTR (detailed illustration: <https://www.dropbox.com/scl/fi/ia0amwbsreujzjil4wp39/assign-the-insert-sequence.dna?rlkey=y50x3pl4pl3nacwsh2giu75tb&dl=0>)

-   So we can claim that "we changed the UG amount in the first 150 bp of the cryptic 3'-UTR"

-   The wild type sequence of "first 150 bp of the cryptic 3'-UTR" has 25 UGs

-   The analysis will looks like:

```{r}
ggplot(df_4 %>% 
         filter(oPools_1 == "wild_type") %>% 
         filter(oPools_3 == "-0%_#1")) + 
  
  geom_point(aes(x = factor(round((first_150_bp_of_cryptic_3_prime_UTR_UG_amount) / 25, 2) * 100 - 100), y = cryptic_fraction * 100, colour = insert, shape = factor(replicate)), position = position_dodge(width = 0.5), alpha = 0.75) + 
  guides(colour = "none") + # this hides the legend for 'shape'
  labs(shape = "replicates") + 
  
  facet_wrap(~Dox_Conc., nrow = 1) + 
    theme_classic() + 
  xlab("change of UG amount (%)") + 
  ylab("cryptic PAS usage (%)") + 
  annotate("segment", x = 8, xend = 8, y = 0, yend = max(df_4$cryptic_fraction) * 100, linetype = "dashed", color = "black", alpha = 0.6) + 
  # the "segment" keyword tells ggplot2 to draw a straight line between two specified points
  theme(text = element_text(size = 15))
```

### Correlation to each oPools

oPools_2

```{r}
ggplot(df_4, aes(x = factor(oPools_2_UG_amount), y = cryptic_fraction, fill = factor(replicate))) + 
  geom_violin() + 
  facet_wrap(~Dox_Conc.) + 
  xlab("UG amount") + 
  geom_point(aes(colour = factor(replicate)), position = position_dodge(width = 1), data = df_4 %>% filter(oPools_2_UG_amount == 27))
```

oPools_1

```{r}
ggplot(df_4, aes(x = factor(oPools_1_UG_amount), y = cryptic_fraction, fill = factor(replicate))) + 
  geom_violin() + 
  facet_wrap(~Dox_Conc.) + 
  xlab("UG amount")
```

oPools_3

```{r}
ggplot(df_4, aes(x = factor(oPools_3_UG_amount), y = cryptic_fraction, fill = factor(replicate))) + 
  geom_violin() + 
  facet_wrap(~Dox_Conc.) + 
  xlab("UG amount") + 
  geom_point(aes(colour = factor(replicate)), position = position_dodge(width = 1), data = df_4 %>% filter(oPools_3_UG_amount == 30))
```

### Overlook

```{r}
ggplot(df_4) + 
  geom_tile(aes(x = factor(Dox_Conc.), y = factor(UG_amount), fill = cryptic_fraction))
```

### Does removing UGs closer to the canonical PAS affect cryptic PAS usage more?

```{r}
df_5 <- df_4 %>% 
  filter(insert == "wild_type---10%_#1---0%_#1" | 
           insert == "wild_type---10%_#2---10%_#2" | 
           insert == "wild_type---10%_#3---10%_#3")

ggplot(df_5, aes(insert, cryptic_fraction)) + 
  geom_point(aes(colour = factor(replicate), shape = insert)) + 
  facet_wrap(~Dox_Conc., nrow = 1) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


df_6 <- df_4 %>% 
  filter(insert == "wild_type---25%_#1---0%_#1" | 
           insert == "wild_type---25%_#2---25%_#2" | 
           insert == "wild_type---25%_#3---25%_#3")

ggplot(df_6, aes(insert, cryptic_fraction)) + 
  geom_point(aes(colour = factor(replicate), shape = insert)) + 
  facet_wrap(~Dox_Conc., nrow = 1) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

df_7 <- df_4 %>% 
  filter(insert == "wild_type---50%_#1---50%_#1" | 
           insert == "wild_type---50%_#2---50%_#2" | 
           insert == "wild_type---50%_#3---50%_#3")

ggplot(df_7, aes(insert, cryptic_fraction)) + 
  geom_point(aes(colour = factor(replicate), shape = insert)) + 
  facet_wrap(~Dox_Conc., nrow = 1) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
